Quantity discrimination, decision-making, and the role of early-life conditions in a lizard

### 2025-07-28 ms.qmd
updates

Quantity discrimination studies have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@mehlis2015quantification; @nieder2018evolution]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in discerning numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) and red-footed turtles (_Chelonoidis carbonaria_) can rapidly learn a quantity discrimination task [@soldati2017long; @lin2021superior; @lin2024trained], and Hermann’s tortoises (_Testudo hermanni_) can discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous]. In lizards, the Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. Shinglebacks (_Tiliqua rugosa_) exhibited spontaneous quantity discrimination using both discrete and continuous cues [@szabo2024spontaneous], while Gidgee skinks (_Egernia stokesii_) discriminated between items when they differed in number but not when size was the main quantitative cue [@szabo_spontaneous_2021]. These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, and that the use of the ANS versus the OFS can differ between species and in different contexts [@szabo_spontaneous_2021]. However, cognitive performance is not solely dictated by species differences; developmental conditions can also shape an individual's ability to process numerical information [@vila_pouca_quantity_2019].

@@

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices of crickets. Specifically, we compared: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. The food consisted of small frozen crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamins. We decided to use frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell]. Lizards were habituated to eating frozen crickets for one month before the experiments.

The crickets used during the experiment were all the same size, but to control for some continuous traits, we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length and total area [see @stancher2015numerical]. We tested if the length occupied by the crickets was similar in both choices (see Supplementary Material - Control of size in both options). Our results showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material - Prey orientation test_).

@@

knitr::include_graphics("./Others/NUMDIS_FIG_1.png")

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets from the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. _Latency_, _Choice_, and _Interest_ were noted as NA when there was no interaction with any of the choices. We also considered _Interest_ as NA in a few cases where the animals managed to lift the dish and eat the prey; _Latency_ and _Choice_ were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were by one observer (PR), who was blind to the treatment of the lizards, but not the type of test or the number of crickets because the videos were analysed manually.

@@

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Estimates of log-latency (Latency), the probability of choosing the larger number of crickets first (Choice), and the estimated interest in the larger number of crickets (Interest) for each of the numerical tests performed across different developmental treatments. The x-axis represents the estimate, and the y-axis is the posterior density of the estimates. The different colours indicate the different treatments. Points and bars represent the median and 95% Highest Density Intervals (95% HPDI) of the estimates, respectively. Vertical dashed lines in Choice and Interest graphs values 0.5 and 0, respectively. Asterisks indicate values significantly different from 0."
#
source(here("R", "func.R"))
#
#### A) Organise the df used for both parts of the figure for each variable
#
# First the df for violin plots for each variable using df_violin_plot function from func.r
df_viol_lat <- df_violin_plot("lat")
df_viol_choice <- df_violin_plot("choice")
df_viol_int <- df_violin_plot("int")
#
# Second the df for the points and bars plot for each variable using df_bars_plot function from func.r
df_bars_lat <- df_bars_plot("lat")
df_bars_choice <- df_bars_plot("choice")
df_bars_int <- df_bars_plot("int")
#
#### B) Make the plots for each variable using plotting function from func.r
#
fig_lat <- plotting("lat")
fig_choice <- plotting("choice") 
fig_int <- plotting("int")
#
#### C) Combine the plots and make final figure
fig_results <- plot_grid(fig_lat, fig_choice, fig_int, ncol = 1) +
  annotate("text", x = 0.305, y = 0.56, label = "*", size = 7, color = "black") + 
  annotate("text", x = 0.61, y = 0.505, label = "*", size = 7, color = "black")
ggsave("./output/figures/fig_results.png", plot=fig_results, width = 25, height = 25, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.png")
```

@@

#### Foraging context may influence spontaneous quantity discrimination in reptiles  
The lack of numerical discrimination in spontaneous tests may be due to _L. guichenoti_ being insectivorous. In studies where reptiles discriminated between different amounts of food, the relevant stimuli involved either vegetables [@gazzola2018continuous; @szabo_spontaneous_2021] or large quantities (e.g., 5 vs. 10) [@recio_prey_2021]. In contrast, when _Podarcis siculus_ were presented with different numbers of _Musca domestica_ larvae in a spontaneous choice test, the lizards failed to discriminate between quantities despite showing quantity discrimination when prey size was the main cue [@miletto2017quantitative] or when tested using training procedures [@miletto2018quantity]. Notably, the numbers of food items employed in @miletto2017quantitative were always fewer than four. Spontaneous choice tests assume that animals should select the most appropriate quantity, which may not necessarily be the largest one [@agrillo2014spontaneous] if the costs of selecting the larger option outweigh the benefits. For example, if capturing larger prey demands more time or energy or success rates remain similar regardless of prey quantity, animals may not prefer larger quantities [@agrillo2014spontaneous]. As a result, even if diet might not affect quantity discrimination abilities, insectivorous lizards might not have evolved a preference for larger patches of prey, at least with small numbers [see @recio_prey_2021].

Previous research showed that other species within the Scincidae family use exclusively the OFS in spontaneous discrimination tests [@szabo_spontaneous_2021]. Here, we aimed to investigate whether _L. guichenoti_ could discriminate between different numbers of crickets in a spontaneous choice test relying exclusively on the OFS. To ensure that lizards decision relied on tracking discrete items rather than using non-numerical cues, we controlled for total length and total area — two variables that could influence ANS processing. We did this by rotating the position of the crickets in the smaller-choice option, making the total length and total area comparable in both choices [@agrillo2014spontaneous; @stancher2015numerical; @cooper2024tell]. The absence of discrimination can therefore be explained by _L. guichenoti_ relying on the ANS for quantity discrimination. Nonetheless, in two of our tests (1 VS 4 and 1 VS 3 crickets) cricket rotation was insufficient to control for total length (see _Supplementary Material - Control of size in both options_). Yet, we still did not find quantity discrimination in _L. guichenoti_, which suggests that neither the OFS nor the ANS were used. Instead, our results likely reflect a genuine lack of quantity discrimination in this species, at least in spontaneous choice tests.

#### Foraging behaviour is robust to early environmental experiences 
Although the effects of early environment on cognition are widely studied, little attention has been paid to its impacts on numerical discrimination and decision-making. We found that decision-making was not influenced by CORT elevations or incubation temperature, as latency to make a choice did not vary across prenatal conditions, which suggests that _L. guichenoti_ may be robust to these early-life factors, at least in the context of foraging decisions. Previous studies have reported that pine snakes (_Pituophis melanoleucus_) incubated at low temperatures were slower and less proficient in capturing prey [@burger_effects_1991], but cold incubated bearded dragons (_Pogona vitticeps_) were faster at approaching a novel object at least at a young age [@siviter_impact_2017]. Maternally transmitted GCs are also known to reduce exploration and neophilia in different taxa [@koolhaas1999coping; @hope_incubation_2018]. Additionally, both prenatal CORT and incubation temperature can influence metabolic rate (MR), which in turn could affect motivation in food-based learning paradigms [@noble_developmental_2018; @cossin2022effect; but see @trnik_persistent_2011; @crino_corticosterone_2014-learn]. In our design, we habituated lizards to the test conditions for two weeks before the experiment to minimise neophilia. And to control for motivation, we standardised hunger levels by providing only one cricket per day and fasting lizards for over 20 hours before testing. Additionally, we excluded trials where lizards failed to interact with either option or refused to eat the cricket provided after testing. These measures likely minimised variation in decision latency due to differences in exploration, neophilia, or hunger levels, which could explain why we did not detect an effect of prenatal conditions. Interestingly, we found that males and older lizards made decisions faster, although these factors were not central to our main hypotheses. However, it is possible that sex and age interact with early-life conditions in shaping behavioral responses. Future research should investigate potential interactions between developmental environment, sex, and age to better understand how these factors influence decision-making in reptiles.

@@

The absence of numerical discrimination in _L. guichenoti_ is consistent with previous studies showing that insectivorous reptiles fail in spontaneous choice tests. However, it is critical to note that the methodologies employed in this study may not have fully captured the nuances of numerical cognition in _L. guichenoti_. Therefore, we recommend that future studies employ a broader array of tests and paradigms to more comprehensively assess numerical competence in this and related species. Furthermore, it would be relevant to study the effect of early environment on quantity discrimination to advance our understanding of numerosity across taxa.

>>>>

Quantity discrimination studies have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@mehlis2015quantification; @nieder2018evolution]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in discerning numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) and red-footed tortoises (_Chelonoidis carbonaria_) can rapidly learn a quantity discrimination task [@soldati2017long; @lin2021superior; @lin2024trained], and Hermann’s tortoises (_Testudo hermanni_) can discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous]. In lizards, the Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. Shinglebacks (_Tiliqua rugosa_) exhibited spontaneous quantity discrimination using both discrete and continuous cues [@szabo2024spontaneous], while Gidgee skinks (_Egernia stokesii_) discriminated between items when they differed in number but not when size was the main quantitative cue [@szabo_spontaneous_2021]. These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, and that the use of the ANS versus the OFS can differ between species and in different contexts [@szabo_spontaneous_2021]. However, cognitive performance is not solely dictated by species differences; developmental conditions can also shape an individual's ability to process numerical information [@vila_pouca_quantity_2019].

@@

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices of crickets. Specifically, we compared: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Each numerical test was made only one per lizard. The food consisted of small thawed frozen crickets (*A. domestica*) that were dusted with calcium and multivitamins. We decided to use thawed frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell]. Lizards were habituated to eating thawed frozen crickets for one month before the experiments.

The crickets used during the experiment were all the same size, but to control for some continuous traits, we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length and total area [see @stancher2015numerical]. We tested if the length occupied by the crickets was similar in both choices (see _Supplementary Material: Control of size in both options_). Our results showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material: Prey orientation test_).

@@

knitr::include_graphics("./Others/NUMDIS_FIG_1.pdf")

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. After the first interaction with any of the options, we recorded the time spent interacting with that option for a maximum of five minutes. We recorded this time for both amounts of food independently of the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time spent interacting with the smaller food option from the time spent interacting with the larger one. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. _Latency_, _Choice_, and _Interest_ were noted as NA when there was no interaction with any of the choices. We also considered _Interest_ as NA in a few cases where the animals managed to lift the dish and eat the prey; _Latency_ and _Choice_ were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were by one observer (PR), who was blind to the treatment of the lizards, but not the type of test or the number of crickets because the videos were analysed manually.

@@

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Estimates of log-latency (Latency), the probability of choosing the larger number of crickets first (Choice), and the estimated interest in the larger number of crickets (Interest) for each of the numerical tests performed across different developmental treatments. The x-axis represents the estimate, and the y-axis is the posterior density of the estimates. The different colours indicate the different treatments. Points and bars represent the median and 95% Highest Density Intervals (95% HPDI) of the estimates, respectively. Vertical dashed lines in Choice and Interest graphs values 0.5 and 0, respectively. Asterisks indicate values significantly different from 0."
#
source(here("R", "func.R"))
#
#### A) Organise the df used for both parts of the figure for each variable
#
# First the df for violin plots for each variable using df_violin_plot function from func.r
df_viol_lat <- df_violin_plot("lat")
df_viol_choice <- df_violin_plot("choice")
df_viol_int <- df_violin_plot("int")
#
# Second the df for the points and bars plot for each variable using df_bars_plot function from func.r
df_bars_lat <- df_bars_plot("lat")
df_bars_choice <- df_bars_plot("choice")
df_bars_int <- df_bars_plot("int")
#
#### B) Make the plots for each variable using plotting function from func.r
#
fig_lat <- plotting("lat")
fig_choice <- plotting("choice") 
fig_int <- plotting("int")
#
#### C) Combine the plots and make final figure
fig_results <- plot_grid(fig_lat, fig_choice, fig_int, ncol = 1) +
  annotate("text", x = 0.305, y = 0.56, label = "*", size = 7, color = "black") + 
  annotate("text", x = 0.61, y = 0.505, label = "*", size = 7, color = "black")
ggsave("./output/figures/fig_results.pdf", plot=fig_results, width = 25, height = 25, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.pdf")
```

@@

#### Foraging context may influence spontaneous quantity discrimination in _L. guichenoti_  
The lack of numerical discrimination in food-related spontaneous tests in _L. guichenoti_ may be due to their insectivorous diet. In studies where reptiles discriminated between different amounts of food, the relevant stimuli involved either vegetables [@gazzola2018continuous; @szabo_spontaneous_2021] or large numbers or alive prey (e.g., 5 vs. 10) [@recio_prey_2021]. In contrast, when _Podarcis siculus_ were presented with different numbers of _Musca domestica_ larvae in a spontaneous choice test, the lizards failed to discriminate between quantities despite showing quantity discrimination when prey size was the main cue [@miletto2017quantitative] or when tested using training procedures [@miletto2018quantity]. Notably, the numbers of food items employed in @miletto2017quantitative were always fewer than four. Spontaneous choice tests assume that animals should select the most appropriate quantity, which may not necessarily be the largest one [@agrillo2014spontaneous] if the costs of selecting the larger option outweigh the benefits. For example, if capturing larger prey demands more time or energy or success rates remain similar regardless of prey quantity, animals may not prefer larger quantities [@agrillo2014spontaneous]. Similarly, animals would not show any preference towards larger number of food items if the likelihood of acquiring one item is the same regardless of quantity [@agrillo2014spontaneous]. As a result, even if diet might not affect quantity discrimination abilities, insectivorous lizards might not have evolved a preference for larger patches of prey, at least with small numbers [see @recio_prey_2021].

Previous research showed that other species within the Scincidae family use exclusively the OFS in spontaneous discrimination tests [@szabo_spontaneous_2021]. Here, we aimed to investigate whether _L. guichenoti_ could discriminate between different numbers of crickets in a spontaneous choice test relying exclusively on the OFS. To ensure that lizards decision relied on tracking discrete items rather than using non-numerical cues, we controlled for total length and total area — two variables that could influence ANS processing. We did this by rotating the position of the crickets in the smaller-choice option, making the total length and total area comparable in both choices [@agrillo2014spontaneous; @stancher2015numerical; @cooper2024tell]. The absence of discrimination in our experiment could be explained by _L. guichenoti_ relying on the ANS for quantity discrimination. Nonetheless, in two of our tests (1 VS 4 and 1 VS 3 crickets) cricket rotation was insufficient to control for total length (see _Supplementary Material: Control of size in both options_). Yet, we still did not find quantity discrimination in _L. guichenoti_, which suggests that neither the OFS nor the ANS were used. Instead, our results likely reflect a genuine lack of quantity discrimination in foraging-related spontaneous choice tests.

#### Foraging behaviour is robust to early environmental experiences 
Although the effects of early environment on cognition are widely studied, little attention has been paid to its impacts on numerical discrimination and decision-making. We found that decision-making was not influenced by CORT elevations or incubation temperature, as latency to make a choice did not vary across prenatal conditions, which suggests that _L. guichenoti_ may be robust to these early-life factors, at least in the context of foraging decisions. Previous studies have reported that pine snakes (_Pituophis melanoleucus_) incubated at low temperatures were slower and less proficient in capturing prey [@burger_effects_1991], and cold incubated bearded dragons (_Pogona vitticeps_) made more errors in a foraging task than warm-incubated lizards [@siviter2019egg]. Maternally transmitted GCs are also known to reduce exploration and neophilia in different taxa [@koolhaas1999coping; @hope_incubation_2018]. Additionally, both prenatal CORT and incubation temperature can influence metabolic rate (MR), which in turn could affect motivation in food-based learning paradigms [@noble_developmental_2018; @cossin2022effect; but see @trnik_persistent_2011; @crino_corticosterone_2014-learn]. In our design, we habituated lizards to the test conditions for two weeks before the experiment to minimise neophilia. And to control for motivation, we standardised hunger levels by providing only one cricket per day and fasting lizards for over 20 hours before testing. Additionally, we excluded trials where lizards failed to interact with either option or refused to eat the cricket provided after testing. These measures likely minimised variation in decision latency due to differences in exploration, neophilia, or hunger levels, which could explain why we did not detect an effect of prenatal conditions. Interestingly, we found that males and older lizards made decisions faster, although these factors were not central to our main hypotheses. However, it is possible that sex and age interact with early-life conditions in shaping behavioral responses. Future research should investigate potential interactions between developmental environment, sex, and age to better understand how these factors influence decision-making in reptiles.

@@

The absence of numerical discrimination in _L. guichenoti_ in food-related spontaneous tests align with previous studies on insectivorous reptiles. However, it is critical to note that the methodologies employed in this study may not have fully captured the nuances of numerical cognition in _L. guichenoti_. Therefore, we recommend that future studies employ a broader array of tests and paradigms to more comprehensively assess numerical competence in this and related species. Furthermore, it would be relevant to study the effect of early environment on quantity discrimination to advance our understanding of numerosity across taxa.
